% !TeX root = ../main.tex

From the day I first met my advisor Tyler Sorensen (on Zoom, while I was still going through chemotherapy, at the very start of the COVID pandemic), he has been nothing but encouraging, empathetic, and focused on my success. Despite coming into my PhD with no background in GPU programming or how to conduct research, Tyler was patient and guided me through the early stages of graduate school as I juggled research, classes, and teaching assistant responsibilities. Tyler always had high standards for my work, but I never felt pressured to work harder or longer than I wanted to, and the quality of the papers that we ended up publishing together reflects his commitment to my growth as a researcher. Your advisor shapes not only your academic career, but also your personal journey during the PhD, and I feel extremely lucky to have had Tyler's support and guidance for all the academic and personal decisions I've made these past six years. From the first time we met in person at a park in Santa Cruz, to attending conferences in Chicago, Vancouver, Seattle, and Singapore, I have enjoyed all the time I've spent working and hanging out with Tyler, and I would not be the person I am today without him. Thank you, Tyler.

I have had the privilege of learning from and being mentored by many others during my PhD. I want to thank Andi Quinn, who became involved in our research group during her first year as a professor at UC Santa Cruz, and who played an especially important role in helping me write my first research paper. I would like to thank Andi and Cormac Flanagan for sitting on both my advancement and defense committees, and John Wickerson for sitting on my advancement committee. Your feedback and support have been invaluable. I also thank Lindsey Kuper and Peter Alvaro, whose online presence and writings led me to apply to UC Santa Cruz to begin with, and from whom I learned a lot during my time as a student in their classes and as a member of the LSD lab. I thank my NDSEG mentor, Rohan Ramlall, for providing an outside voice and perspective for the past three years, and Randy Rauwendaal, who has given me valuable career opportunities and advice. I would also like to thank my Apple internship mentors Justin Hensley and Davoud Jamshidi, who helped me navigate research in industry for the first time. I am deeply grateful to Bill Collier, who as one of the leaders in memory consistency wrote the seminal book \textit{Reasoning About Parallel Architectures} in 1992, but still engaged thoughtfully with my work as a new graduate student 30 years later. Sadly, Bill passed away in late 2025, but I am indebted to him and all the other pioneers in this field.

During the past six years, I have had the fortune of collaborating with many researchers outside UC Santa Cruz, including Vijay Nagarajan, Soham Chakraborty, Andr\'es Goens, Susmit Sarkar, Dennis Sprokholt, Soham Bagchi, Hern\'an Ponce de Le\'on, Raph Levien, David Neto, Alan Baker, Cheng Gu, Zhenkai Zhang, and Yanan Guo. Thank you for all the engaging discussions, and I'm proud of the research and publications we've achieved together.

At UC Santa Cruz, I had the fortune of working with many other amazing students. To Yanwen Xu, thank you for your friendship as we navigated being Tyler's first two PhD students, and congrats on making it out of the PhD before me. To Rithik Sharma, Jessica Dagostini, Zheyuan Chen, and Devon McKee, thank you for the times spent working together, giving feedback on each other's presentations, and all the time spent hanging out at group dinners. Thank you also to the amazing undergrads and master's students who I have worked with and mentored at UC Santa Cruz: Tianhao (Tim) Guo, Mingun Cho, Kyle Little, Ashley Lee, Sean Siddens, Sanya Srivastava, Gurpreet Dhillon, James Contini, and Neha Abbas. You have all taught me as much if not more than I taught you.

To my parents, Marc Levine and Elaine Ebner, thank you for the life you've given me. You always encouraged me to do my best, but never put pressure on me to do more than I could, and your support in everything I've done and gone through has played a large part in all my successes. To my sister, Delaney, thank you for all the memories running around our backyard and the creek in Orinda, and I'm looking forward to seeing you graduate with your master's from UCLA next year.

Thank you to all my friends and all those who I've spent time with on the bike, in the mountains, or on the water these past six years. This thesis represents the culmination of my academic achievements, but my life outside of research has given me the balance I need to do my best work.

And finally, to Sarah. There is so much I could say about the ten years we've spent together, the life we've been building together. Since I first decided to do my PhD, our life has been full: adopting Hunter, getting engaged, getting married, trips all over the country and the world. Through it all, you have been a steadying and motivating partner, taking care of me at my lows, sharing the joys of all my highs. My success is yours, and I'm so excited to see what the next chapter of our lives has in store for us. I love you.
