% !TeX root = ../main.tex
\chapter[Conclusion]{Conclusion}
\label{chap:conclusion}

\epigraph{\textit{``Because after the finish all the suffering turns to memories of pleasure, and the greater the suffering, the greater the pleasure.''}}{--- Tim Krabbe, \\\footnotesize\textit{The Rider}}

The aim of this thesis was to develop rigorous techniques for validating the correctness of GPU memory consistency specification (MCS) implementations and the memory safety of complex GPU frameworks. This required addressing three related challenges: evaluating whether an MCS testing environment is effective, scaling testing across diverse GPU systems, and understanding how memory consistency interacts with memory safety in WebGPU. The resulting techniques expose bugs that previous approaches missed, enable testing across many GPUs, and identify safety risks created by data races across layered GPU compilation stacks.

MC Mutants establishes a methodology, grounded in formal models and empirical testing techniques, for evaluating and improving testing environments. GPUHarbor applies scalable and portable testing to characterize MCS behavior across a wide range of devices and to support deeper investigations of GPU memory systems. SafeRace develops new tools to evaluate memory safety, identifies a specification vulnerability in WebGPU, and proposes guarantees that preserve safety in the presence of data races. This chapter summarizes these contributions and discusses directions for future work.

\section{Contributions}

This thesis makes the following original contributions:

\begin{enumerate}
    \item \textbf{MC Mutants: Evaluating and Improving Testing for Memory Consistency Specifications} (\mychap\ref{chap:mc-mutants}). This work introduces MC Mutants, a mutation-testing methodology that quantitatively evaluates MCS testing environments, along with parallel testing and MCS test confidence strategies. Across four GPUs from four vendors, parallel testing improves mutant death rates by an average of three orders of magnitude and enables high-confidence testing with a time budget of approximately one minute per test. The evaluation identifies two WebGPU MCS bugs, contributes tests adopted by the WebGPU conformance test suite, and leads to an AMD driver fix and a WebGPU specification change. This work appeared at ASPLOS 2023 and received both Distinguished Paper and Distinguished Artifact awards.

    \item \textbf{GPUHarbor: Testing GPU Memory Consistency at Large} (\mychap\ref{chap:gpuharbor}). This work introduces GPUHarbor, a portable MCS testing suite with accessible web and Android interfaces, and uses it to conduct a large-scale study of GPU weak memory behaviors. The study collects results from 106 devices spanning seven vendors, executes nearly 400 billion test iterations, and identifies two previously unreported MCS bugs on Arm and NVIDIA devices. The testing techniques also support detailed investigations of modern NVIDIA GPU memory consistency and demonstrate that weak memory behaviors can form the basis of GPU side-channels. GPUHarbor appeared at ISSTA 2023 and received a Distinguished Artifact award, and its techniques were later used in several follow-on publications about NVIDIA GPU MCSs and GPU side-channels.

    \item \textbf{SafeRace: Assessing and Addressing WebGPU Memory Safety in the Presence of Data Races} (\mychap\ref{chap:saferace}). This work identifies a data race specification vulnerability caused by the mismatch between WGSL's memory-safety guarantees and the undefined behavior permitted by its compilation targets. Bottom-up and top-down threat assessments discover memory isolation vulnerabilities affecting AMD and Apple GPUs and a pre-release version of Firefox. To address the specification vulnerability, SafeRace introduces the SafeRace Memory Safety Guarantee, implements its first property as a WGSL compiler pass, and empirically evaluates its second property across existing WebGPU implementations. This work appeared at OOPSLA 2025 and provides a framework for reasoning about safety across compositional language and compilation designs.
\end{enumerate}

\section{Future Work \label{sec:conclusion-future-work}}

The research presented in this thesis is a stepping stone toward many other impactful projects, some of which are detailed below.

\paragraph{Memory Consistency Testing.} The techniques presented here show that MCS testing can be evaluated more rigorously and scaled in ways that allow large datasets to be collected and new frameworks and devices to be tested relatively quickly. One area where new empirical testing methods are still needed is heterogeneous environments, i.e., environments where multiple processors share cache-coherent memory, such as the Grace-Hopper Superchip. Some work has started to explore this domain both formally~\cite{goens2023compound} and empirically~\cite{srivastava2024testing}, but more research is needed to develop testing strategies that effectively cover the full breadth of heterogeneous MCSs. Beyond CPUs and GPUs, domain-specific accelerators like TPUs, FPGAs, and other neural engines continue to grow in popularity but lack publicly available MCSs or empirical studies that analyze their shared-memory behaviors.

\paragraph{Probabilistic Memory Consistency Specifications.} Approximate-computing applications can sometimes trade occasional inaccurate results for increased performance by relaxing synchronization requirements~\cite{hogwild,10.1145/2540708.2540711}. A similar tradeoff may be possible for memory consistency. Although an MCS defines which weak behaviors are allowed, the GPUHarbor results in \mychap\ref{chap:gpuharbor} demonstrate that the rates at which these behaviors occur vary substantially across devices and system conditions. This observation motivates \textit{probabilistic memory consistency specifications} (P-MCSs), which would augment an MCS with empirically derived estimates of weak-behavior rates. Comprehensive litmus-test campaigns could be used to synthesize device-specific P-MCSs and evaluate conditions under which approximate-computing applications might omit expensive synchronization while satisfying probabilistic accuracy requirements. Unlike relaxed-synchronization techniques designed for particular algorithms, P-MCSs would provide reusable information for reasoning about accuracy and performance tradeoffs across systems.

\paragraph{Memory Safety and Security.} The work in \mychap\ref{chap:saferace} is an early attempt to improve the memory safety and security of GPU programming languages that require more robust guarantees. However, as more sensitive computation continues to be offloaded to GPUs, e.g., running large language models, frameworks and runtimes will need to be hardened further. WebGPU attempts to provide safety through a combination of transformations during compilation and lower-level framework guarantees, but prototype GPU languages have also been developed that provide memory safety by default~\cite{kopcke2024descend}. New research can build on these techniques to define memory-safe languages that avoid performance loss and to explicitly model and test complex GPU compilation stacks, providing a cross-platform target for GPU computation.

This thesis has also shown that MCSs present a vector for creating new hardware side-channels for GPUs. While side-channels based purely on observing weak behaviors are not high-capacity, the work in this thesis has already been used as a foundation to create a more sophisticated side-channel attack on NVIDIA MIG technology. The hardware mechanisms behind weak behaviors are complex and include many hardware components, so it may be fruitful to conduct more extensive experiments and construct a taxonomy of different litmus-test behaviors on various hardware. These results could then provide a foundation for constructing targeted side-channels after the initial MCS-testing-based feasibility stage.

\paragraph{Novel Cross-Platform GPU Applications.} The underlying goal of validating the memory consistency and safety of GPU frameworks is to enable the development of correct and performant applications using those frameworks. For example, one project not discussed in this thesis was my development of a WebGPU backend for llama.cpp, allowing local LLMs to run directly in the browser with GPU acceleration. This project improved on other browser-based LLM inference engines by being more memory-efficient, developing portable shaders that run efficiently across GPUs from many vendors with diverse hardware characteristics, and supporting many weight precisions developed by the machine-learning research community. This research would not have been possible without the validation work presented in this thesis and lays the foundation for future development of cross-platform GPU applications.

\section{Final Perspective}

For the foreseeable future, GPUs will continue to play a large role in computing applications. Of course, the rise of AI has led to a boom in the fortunes of GPU chipmakers, especially for data-center GPUs used to train and serve the largest language models, but GPUs also power many other machine-learning, image-processing, and scientific workloads. Bringing this power to all applications, whether they are running on large discrete GPUs or power-efficient mobile GPUs, motivates the development of cross-platform and user-friendly frameworks like WebGPU. The work in this thesis contributes to the reliability and safety of these frameworks and will hopefully continue to have a direct impact and inspire future research in these areas.

More fundamentally, however, this thesis speaks to the idea that formal models should be more directly connected to actual implementations through empirical methodologies like the ones described here. Much of the time, formal models for programming frameworks are defined by committees that have an idea of how systems should behave but rely on informal descriptions or incomplete understandings of how the systems are actually implemented. As shown by this thesis, mismatches between formal models and implementations can lead to compiler bugs, security vulnerabilities, and confusion about how systems behave. I don't think there are any easy answers here, and testing will never provide a complete guarantee, but this thesis takes concrete steps toward connecting models with empirical observations, and the lessons learned will continue to guide my work for the rest of my career.

Finally, memory consistency is just one realization of the broader field of consistency, and GPUs are just one type of parallel processor. In fact, early in my PhD I was also interested in distributed-systems research, since that was my background prior to returning to graduate school. Alas, time is limited, but there are still deep connections and a shared history between distributed-systems consistency and shared-memory consistency, as well as with any system where multiple components maintain their own views of data and share them with others. Future chips and distributed systems may not look anything like today's GPUs, but they will still likely have to grapple with the problem of consistency, and I am proud that this thesis contributes in some small way toward our understanding of this domain.

\paragraph{A Note on AI.} It is not lost on me, nor likely on anyone reading this thesis, that at the time of writing we are in the midst of a technological revolution. Large language models, which in large part are trained and served on some of the same GPUs on which this thesis runs memory consistency tests, can now write code and design complex systems in ways that seemed exceedingly futuristic even when I started my PhD. In some ways, I regret not having access to them earlier, as some of the work in this thesis would likely have progressed much more quickly with their help. However, in other ways, I am glad I had the chance to complete the main portions of my thesis prior to their arrival, as the time I spent poring over documentation and unfamiliar codebases was a learning experience I highly value.

Nevertheless, in the words of Bob Dylan, ``the times they are a-changin','' and even now, when I'm coding I spend most of my time conversing with an agent in natural language rather than typing out the syntax myself. So far, AI has enabled work, especially technical work, to be done much more quickly, but in my experience, the tools still require expert guidance about what directions to take the work, i.e., research taste~\cite{olah_research_taste}. Model capabilities continue to progress quickly, however, and it is not yet clear what the true limits of current model-building techniques are.

In research, AI is changing not only what we should work on, but how research itself should be done~\cite{cheng2025barbariansgateaiupending}. My take on this, at least so far, is that we should continue to work on the hard problems that we always have. There is still room in today's world for good ideas, and good ideas will continue to lead to real improvements in the systems to which they are applied. Perhaps the bigger question is how good ideas will shine through in a world filled with AI-generated output; for example, a leading AI conference recently rejected 18.4\% of submissions to its position-paper track for being clearly AI-generated~\cite{neurips2026_ai_generated_papers}, while 21\% of paper reviews for another leading AI conference were estimated to be AI-generated~\cite{pangram2025_iclr_reviews_ai}. For any up-and-coming researchers reading this, it's hopefully needless to say, but don't do this.

I guess it might be good to disclose the ways I used AI to help write this thesis. First, every word has been reviewed and edited by me, and I take responsibility for all the results. AI tools were used to help format figures, improve layout, and provide feedback on spelling, grammar, clarity, and continuity. In some cases, AI was used to generate an initial draft of a paragraph or short section based on an outline I wrote, which I then edited. Much of the technical content was, for better or worse, drawn from papers written before AI tools were available. Hopefully, this thesis is free of any ``AI-isms,'' e.g., ``it's not X, it's Y,'' ``same X, same Y,'' ``why this matters,'' liberal use of em-dashes---like this---and colon sentence structures that relate two clauses: because, according to AI, the colon is a powerful grammatical tool.

\paragraph{Fin.} Despite the quote at the beginning of this chapter, there was not too much suffering involved in either doing the work or writing this thesis. I appreciated every moment of the PhD, the highs and the lows, the support of my advisor and everyone else in my life, the time spent doing research and avoiding it. And now, it is done, and there's likely a small summer south swell pushing into the reef down on East Cliff, and the surf is calling.

\epigraph{\textit{``I continued to doubt. But I was not afraid. I just didn't want this to end.''}}{--- William Finnegan, \\\footnotesize\textit{Barbarian Days: A Surfing Life}}
